---
title: "Spillover of Disease Control Costs In the Presence of Interference"
subtitle: "Network Estimation with Maximum Likelihood"
author:   
  - name: Trent VanHawkins
    affiliations:
      - ref: osu
    degrees: MS
    corresponding: true
  - name: Jaeyoung Hwang
    affiliations:
      - ref: usda
    degrees: PhD
  - name: David Gent
    affiliations:
      - ref: usda
      - ref: osu
    degrees: PhD
  - name: Shirshendu Chatterjee
    affiliations:
      - ref: nyu
    degrees: PhD
  - name: Sharmodeep Bhattacharyya
    affiliations:
      - ref: osu
    degrees: PhD
affiliations:
  - id: osu
    name: Oregon State University
  - id: usda
    name: U.S. Department of Agriculture
  - id: nyu
    name: City University of New York
format: 
  pdf:
    geometry: "margin=0.75in"
    mathspec: true
    cap-location: bottom
    tbl-cap-location: top
    header-includes: 
      - \usepackage{fancyhdr, amsthm, amssymb,amsfonts,amsthm, amsmath, bbm, tikz}
      - \usepackage{float, tabularx}
      - \usetikzlibrary{positioning, arrows.meta} 
      - \floatplacement{table}{H}
      - \pagestyle{fancy}
      - \fancyhead[R]{Model Considerations}
      - \fancyhead[L]{Trent VanHawkins}
      - \fancyfoot[C]{\thepage} # Center page number at bottom of each page
page-layout: full
bibliography: references.bib
---

# Introduction

Hop Powdery Mildew *Podosphaera macularis* has been the focus of much of the work in this research group. The group has previously investigated spread and control of the disease via Monte-Carlo simulations [@pedro] and have recently taken the first steps in estimating spillover effects of pesticide use on neighboring hop-yards (Hwang et. al., 202X). Estimation of the underlying disease transmission network is foundational to these works, and has previously proven to be a complex exercise in non-linear optimization. This complexity is, in part, introduced by a pathological constraint. Put simply, Hop cultivars can be dichotomized on the basis of resistance to races of *P. macularis*. Cultivars possessing the R6 resistance gene can infected *only by the V6 race* of *P. macularis*, while the remaining cultivars may be infected by other races (Vb, V3, V4, V5) *and* the V6 race (@fig-transmission-graph). 

::: {#fig-transmission-graph}
\begin{tikzpicture}[
    node style/.style={circle, draw=black, minimum size=22mm, align=center, font=\small\bfseries, thick},
    edge style/.style={thick, >=Stealth},
    v6_style/.style={fill=orange!20},
    non_v6_style/.style={fill=blue!10}
]

    \node[node style, v6_style]     (R6)   at (0, 0)   {R6};
    \node[node style, non_v6_style] (NR6)  at (4, 0)   {Non-R6};

    % V6 strain: bidirectional arrow between R6 and Non-R6
    \draw[edge style, <->, orange!70!black]
        (R6) to[bend left=0]
        node[midway, above, font=\scriptsize, orange!70!black] {V6}
        (NR6);

    % Non-V6 strain: self-loop on Non-R6
    \draw[edge style, ->, blue!60!black]
        (NR6) to[out=330, in=30, looseness=3]
        node[right, font=\scriptsize, blue!60!black] {Non-V6}
        (NR6);

    % V6 strain: self-loop on R6
    \draw[edge style, ->, orange!70!black]
        (R6) to[out=210, in=150, looseness=3]
        node[left, font=\scriptsize, orange!70!black] {V6}
        (R6);

    % Legend
    \begin{scope}[shift={(8, 0)}]
        \draw[black, thin, dashed] (-0.3, 1.2) rectangle (2.8, -0.3);
        \node[anchor=west, font=\small\itshape] at (-0.2, 0.9) {Initial strain:};
        \draw[orange!70!black, thick, <->] (0, 0.45) -- (0.6, 0.45);
        \node[anchor=west, font=\small] at (0.7, 0.45) {V6};
        \draw[blue!60!black, thick, ->]   (0, 0.05) -- (0.6, 0.05);
        \node[anchor=west, font=\small] at (0.7, 0.05) {Non-V6};
    \end{scope}

\end{tikzpicture}

\small A simplified graph of possible network transmission between yards, conditioning on
cultivar type and initial strain. R6 and non-R6 yards transmit to each other under a V6
initial strain, while non-R6 yards under a non-V6 strain can only infect other non-R6 yards
(self-loop).
:::

Information provided by the sentinel dataset on cultivar type and initial strain allow us to apply the contraint in our modeling procedure. Up to now, all approaches of this group have taken the approach of @gent2019, in which estimation is conducted on discrete time transitions seperately for R6 and non-R6 yards (@eq-old-mod). The resulting directed edge-weights of the transmission network are then estimated accordingly (@eq-old-edge-weights). The advantage of this approach is that it allows flexible estimation of auto-infection and dispersal parameters in the case that they differ between R6 and non-R6 cultivars. The potentail drawback is that less data is used to estimate each set of parameters potentially reducing statistical power and parameter identifiability---particularly during early months (April-May) when there are relatively few infections. 

$$
\log\left(\frac{p_{i,t}}{1-p_{i,t}}\right) = \beta + \delta\underbrace{\left(\frac{y_{i, t-1}}{n_{i, t-1}}\cdot \exp(-\eta_1 s_{i,t-1})\right)}_{\text{auto-infection}} + \gamma \underbrace{\sum_{j\neq i} \left(\frac{a_j \cdot y_{j,t-1}}{n_{j,t-1}} \cdot \exp(-\eta_2 s_{j, t-1}) \cdot w_{i,j} \cdot (1 + d_{i,j})^{-\alpha} \right)}_{\text{recieved inoculum}}
$$ {#eq-old-mod}

$$
W_{ij} = \begin{cases} 
\gamma_1 \frac{a_j \cdot y_{j,t-1}}{n_{j,t-1}} \cdot \exp(-\eta_{21} s_{j, t-1}) \cdot w_{i,j} \cdot (1 + d_{i,j})^{-\alpha_1} \cdot \mathbf{1}_{V_6}(j) & \text{Yard $i$ is R6} \\
\gamma_2 \frac{a_j \cdot y_{j,t-1}}{n_{j,t-1}} \cdot \exp(-\eta_{22} s_{j, t-1}) \cdot w_{i,j} \cdot (1 + d_{i,j})^{-\alpha_2} & \text{Yard $i$ is non-R6}
\end{cases}
$$ {#eq-old-edge-weights}

::: {#tbl-model-specs}
\begin{table}
    \small
    \centering
    \subcaptionbox{Covariate notations and descriptions.\label{tbl-covariates}}{
        \begin{tabularx}{\textwidth}{lX}
            \toprule
            Covariate & Description \\
            \midrule
            $y_{i, t-1}$     & Number of diseased plants at yard $i$ in the prior month \\
            $n_{i, t-1}$ & Number of plants sampled at yard $i$ in the prior month \\
            $a_i$             & Area (hectares) of yard $i$ \\
            $d_{ij}$          & Distance (kilometers) from centroids of yard $i$ to yard $j$ \\
            $w_{ij}$          & Wind vector on $j - i$ direction in the prior month \\
            $s_i$             & Number of fungicide sprays in yard $i$ in prior month \\
            \bottomrule
        \end{tabularx}
    }
    
    \vspace{1em}  

    \subcaptionbox{Model parameters and descriptions.\label{tbl-parameters}}{
        \begin{tabularx}{\textwidth}{lX}
            \toprule
            Parameter & Description \\
            \midrule
            $\beta$    & Baseline log-odds of disease, after accounting for autoinfection and dispersal. \\
            $\delta$   & Change in log-odds of disease associated with autoinfection at the yard scale, after accounting for disease spread. \\
            $\gamma$   & Distance-adjusted change in log-odds of disease associated with disease spread from other yards, after accounting for autoinfection. \\
            $\alpha$   & Dispersal parameter providing distance adjustment to change in log-odds of disease associated with individual source yards. \\
            $\eta_1$   & Change in log-odds of disease associated with autoinfection at the yard scale, after accounting for fungicide sprays. \\
            $\eta_2$   & Dispersal parameter providing fungicide spray adjustment to change in log-odds of disease associated with individual source yards. \\
            \bottomrule
        \end{tabularx}
    }
\end{table}

\small Descriptions of model parameters and covariates notations for each yard of interest $i = 1,2,...,N$ in a month, and all other yards $j = 1,2, \ldots, N_{-i}$ relative to yard $i$.
:::

Recently, an exploration of multi-layered networks as a theoretical framework for estimation of spillover effects has lead us to an approach that may collapse estimation across cultivars by introducing an indicator function which controls the pathological constraint of @fig-transmission-graph within the model component accounting for incoming infection from other yards. We hypothesize that the phyisological mechanisms responsible for auto-infection and recieved inoculum *do not* differ among cultivar types and that collapsing the framework of @gent2019 into a single regression model will improve parameter identifiability. In the present report, we formally introduce the proposed adaptation and explore its predictive properties empirically.

We additionally note the unique complexity of optimization under the framework of @gent2019, regardless of the specific pathosystem. As mentioned previously, disease networks tend to be sparse during the beginning of an epidemic, often yielding non-identifiable parameters under @eq-old-mod. It is also the case that the form of the auto-infection and recieved inoculum components may vary depending on the pathosystem. For example, we implement a power-law kernel to allow for the "fat-tailed" dispersal gradient often found in Long-Distance Dispersal (LDD) pathosystems; however, the exponential kernel is another common choice. In some cases, it may even be common practice to find the kernel that best fits the data via model selection. Under any optimization scheme which requires gradient information or higher-order derivaties (BFGS, SQLS, etc...) the gradient must be derived (and programmed) for each form of the model under consideration to improve convergence. This process is time-consuming and may limit efforts at finding a best fit. 

The Stan programming language implements **algorithmic (or automatic) differentiation** (AD), which computes derivatives that are exact to machine precision. This eliminates the significant overhead of deriving closed-form expressions and minimizes the risk of manual programmatic errors. While Stan is primarily known for Bayesian inference via Markov Chain Monte Carlo (MCMC), libraries such as `cmdstanr` and `cmdstanpy` provide access to high-performance optimization algorithms—specifically the Limited-memory Broyden–Fletcher–Goldfarb–Shanno (**L-BFGS**) algorithm. These tools allow for a frequentist implementation of the model, yielding Maximum Likelihood Estimates (MLEs) and standard errors derived from the inverse Hessian, while benefiting from Stan’s robust computational architecture. We implement the present model in the Stan programming language in hopes of introducing a flexible optimization routine for future implementations of this work. 

# Methods

## The adapted model 

To adapt @eq-old-mod, we begin by defining indicators (@eq-indicators) that will help us impose the constraint in @fig-transmission-graph. 

$$
c = \begin{cases} 1 & \text{cultivar} = \text{R6} \\
0 & \text{cultivar} = \text{non-R6}
\end{cases} \qquad v = \begin{cases} 1 & \text{strain} = \text{V6} \\
0 & \text{strain} = \text{non-V6}
\end{cases}
$${#eq-indicators}

We can then define our directed adjacency matrix as a function of the indicators in @eq-indicators as in @eq-directed-adj.

$$
\mathbf{A}_{ij} = \begin{cases}
1 & v_j = 1 \text{ or } (c_i = c_j = 0) \\
0 & \text{else}
\end{cases}
$${#eq-directed-adj}

Note that our contsraint is satisfied in @eq-directed-adj if source yard $j$ has initial strain V6 (can infect any other yard) *or* if both source yard $j$ and target yard $i$ have the non-R6 cultivar (can be infected by all races of *P. macularis*). We can then define our new model as @eq-new-mod. 

$$
\log\left(\frac{p_{i,t}}{1-p_{i,t}}\right) = \beta + \delta\left(\frac{y_{i, t-1}}{n_{i, t-1}}\cdot \exp(-\eta_1 s_{i,t-1})\right) + \gamma \sum_{j\neq i} \left(A_{ij} \cdot \frac{a_j \cdot y_{j,t-1}}{n_{j,t-1}} \cdot \exp(-\eta_2 s_{j, t-1}) \cdot w_{i,j} \cdot (1 + d_{i,j})^{-\alpha} \right)
$$ {#eq-new-mod}

This simple modification means that the recieved inoculum component will take on a value of zero if the constraint of @fig-transmission-graph is not met. We can then adapt @eq-old-edge-weights analogously as @eq-new-edge-weights.

$$
W_{ij} = 
\gamma  \cdot A_{ij} \cdot \frac{a_j \cdot y_{j,t-1}}{n_{j,t-1}} \cdot \exp(-\eta_{2} s_{j, t-1}) \cdot w_{i,j} \cdot (1 + d_{i,j})^{-\alpha} 
$$ {#eq-new-edge-weights}

Note that we drop the additional subscript (1 or 2) for whether the parameters are estimated from non-R6 yards or R6 yards. 

## Optimization in Stan
Stan is naturally a Probabalistic Programming Language (ProbProg), and as such requires *prior* specification for each model prameter (or assumes flat priors). For the present dataset, we have potential problems of identifiability and as such, we can leverage these priors regularize our model estimates away from extreme values. This is particularly useful in the case that parameters are maximized "on the boundary" of the parameter space. Stan also allows us to place *constraints* on parameters and then optimizes them in the unconstrained space automatically (e.g. the log-sapce for `param<lower=0>`). 

::: {#tbl-priors}
\begin{table}
\centering
\vspace{0.3cm}
\footnotesize
\begin{tabular}{l c l cc cc}
\toprule
 &  &  & \multicolumn{2}{c}{\textbf{Regularized}} & \multicolumn{2}{c}{\textbf{Informed}} \\
\cmidrule(lr){4-5} \cmidrule(lr){6-7}
\textbf{Parameter}& \textbf{Constraint} & \textbf{$p(\theta)$}& \textbf{Location} ($\mu$) & \textbf{Scale} ($\sigma^2$) & \textbf{Location} ($\mu$) & \textbf{Scale} ($\sigma^2$) \\
\midrule
$\beta$ (Baseline Intercept) & $(-\infty, \infty)$ & Normal & 0.0 & 100.0 & $-2.5$ & 100.0 \\
\addlinespace
$\delta$ (Auto-infection) & $[0, \infty)$ & Log-Normal & 0.0 & 1.0 & 1.0 & 1.0 \\
\addlinespace
$\gamma$ (Dispersal Magnitude) & $[0, \infty)$ & Log-Normal & 0.0 & 1.0 & 7.5 & 1.0 \\
\addlinespace
$\alpha$ (Kernel Decay) & $[0, \infty)$ & Log-Normal & 1.0 & 1.0 & 1.0 & 1.0 \\
\addlinespace
$\eta_1$ (Auto-infection Decay) & $[0, \infty)$ & Log-Normal & 0.0 & 1.0 & 0.0 & 1.0 \\
\addlinespace
$\eta_2$ (Dispersal Decay) & $[0, \infty)$ & Log-Normal & 0.0 & 1.0 & 0.0 & 1.0 \\
\bottomrule
\end{tabular}
\end{table}

\small Parameter Constraints and Hyperparameter Distributions under Alternative Prior Scenarios
:::

Note from @tbl-priors, we tried two different types of priors. One set that simply regularizes towards zero (closely related to the L2 penalty) and one that uses information from previous maximization attempts (informed). Due to the non-linear nature of the model and potention multi-modality, we tried 10 sets of randomly generated initial values from the priors and checked for convergence to a global maximum. 

# Results
Below we present the parameter estimates and standard errors under each prior scenario presented in @tbl-priors. We then conduct some basic forms of model-checking to verify model fit.

## Maximum Likelihood Estiamtes

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(kableExtra)

mle_df <- read.csv(here("data/processed/results/mle_results.csv"))
mle_clean <- mle_df |> select(month, scenario, parameter, n_converged, estimate, std_error)

generate_latex_mle_table <- function(df, output_path = NULL) {
  # 1. Clean and normalize values
  df_clean <- df %>%
    mutate(
      month = str_to_lower(month),
      scenario = str_to_lower(scenario),
      parameter = str_to_lower(parameter)
    )
  
  # 2. Average over seeds if multiple exist (just in case)
  df_agg <- df_clean %>%
    group_by(month, scenario, parameter) %>%
    summarise(
      estimate = mean(estimate, na.rm = TRUE),
      std_error = mean(std_error, na.rm = TRUE),
      .groups = "drop"
    )
  
  # 3. Categorical sorting configurations
  month_order <- c("apr", "may", "jun", "jul")
  param_order <- c("beta", "delta", "gamma", "alpha", "eta1", "eta2")
  
  df_sorted <- df_agg %>%
    mutate(
      month = factor(month, levels = month_order, ordered = TRUE),
      parameter = factor(parameter, levels = param_order, ordered = TRUE)
    ) %>%
    filter(!is.na(month), !is.na(parameter)) %>%
    arrange(month, parameter)
  
  # 4. Pivot the data to get scenarios side-by-side
  pivot_df <- df_sorted %>%
    pivot_wider(
      names_from = scenario,
      values_from = c(estimate, std_error),
      names_glue = "{scenario}_{.value}"
    )
  
  # Define parameter LaTeX mapping for clean display
  param_mapping <- c(
    'beta' = "$\\beta$",
    'delta' = "$\\delta$",
    'gamma' = "$\\gamma$",
    'alpha' = "$\\alpha$",
    'eta1' = "$\\eta_1$",
    'eta2' = "$\\eta_2$"
  )
  
  # 5. Format and prepare for kable
  table_data <- pivot_df %>%
    group_by(month) %>%
    mutate(
      # Only show the month label on the first row of each month group
      month_col = if_else(row_number() == 1, str_to_title(as.character(month)), ""),
      parameter_col = param_mapping[as.character(parameter)]
    ) %>%
    ungroup() %>%
    select(
      month_col,
      parameter_col,
      towards_zero_estimate,
      towards_zero_std_error,
      informed_priors_estimate,
      informed_priors_std_error
    ) %>%
    mutate(across(where(is.numeric), ~ sprintf("%.3f", .)))
  
  # Calculate grouping sizes to add space (\addlinespace) dynamically between months
  group_sizes <- pivot_df %>%
    group_by(month) %>%
    summarize(n = n(), .groups = "drop")
  
  end_rows <- cumsum(group_sizes$n)
  space_rows <- end_rows[-length(end_rows)] # Exclude the very last row
  
  # 6. Generate LaTeX table using kableExtra
  latex_table <- kable(
    table_data,
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    align = "ll|cc|cc",
    col.names = c("Month", "Parameter", "Estimate", "Std. Error", "Estimate", "Std. Error"),
    caption = "Maximum Likelihood Estimates (MLE) and Standard Errors Across Scenarios",
    label = "mle-results"
  ) %>%
    kable_styling(latex_options = c("hold_position", "striped"), font_size = 9) %>%
    add_header_above(c(" " = 1, " " = 1, "Regularized" = 2, "Informed" = 2))
  
  # Apply dynamic spacing between month groups
  for (r in space_rows) {
    latex_table <- latex_table %>%
      row_spec(r, extra_latex_after = "\\addlinespace")
  }
  
  # Print and optionally save the LaTeX output
  if (!is.null(output_path)) {
    writeLines(as.character(latex_table), output_path)
    message(paste("\n[Success] Table saved to:", output_path))
  }
  
  return(latex_table)
}
generate_latex_mle_table(mle_clean)
```

Diagnostics reveal improving model fit as transitions progress (@fig-fitted-vs-true). @fig-residuals gives the deviance of predicted points across fitted values. We observe relatively large deviance across the domain of predictions, but they are particularly large at small values, indicating that the model has the hardes time predicting lower disease prevalence. 

```{r}
#| fig-height: 6
#| fig-width: 8
#| label: fig-fitted-vs-true
#| fig-cap: Fitted vs. Observed values
#| echo: false
#| message: false
mle_preds <- read.csv(here("data/processed/results/mle_preds.csv"))

mle_preds |> 
  mutate(month = factor(month, levels = c("may", "jun", "jul"), labels = c("April-May", "May-June", "June-July")),
  scenario = factor(scenario, levels = c("towards_zero", "informed_priors"), labels = c("Regularized", "Informed")),
  year = factor(year, ordered = T)) |> 
  ggplot(aes(y = true_prob, x = pred_prob)) +
  geom_point(aes(color = year))+
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red")+
  facet_grid(month ~ scenario)+
  labs(y = "Observed Proportion (Y/N)", x = "Predicted Prob.", color = "Year") + 
  theme_bw() + 
  theme(legend.position = "bottom")

```

```{r}
#| fig-height: 6
#| fig-width: 8
#| label: fig-residuals
#| fig-cap: Deviance residuals across transitions and prior scenarios
#| echo: false
#| message: false

mle_preds |> 
  mutate(month = factor(month, levels = c("may", "jun", "jul"), labels = c("April-May", "May-June", "June-July")),
  scenario = factor(scenario, levels = c("towards_zero", "informed_priors"), labels = c("Regularized", "Informed")),
  year = factor(year, ordered = T)) |> 
  ggplot(aes(y = deviance_resid, x = pred_prob)) +
  geom_point(aes(color = year))+
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")+
  facet_grid(month ~ scenario)+
  labs(y = "Deviance", x = "Predicted Prob.", color = "Year") + 
  theme_bw() + 
  theme(legend.position = "bottom")

```

# Discussion
